---
title: "Class 3"
author: "Colin Kuehl"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

Today we'll start by by looking at data by Bush and Clayton on gender and climate. 



```{r}
#install.packages("tidyverse") ## This package is used to tidydata. We'll use it later
#install.packages("ggplot2") #Very popular graphing package

library(tidyverse)
library(ggplot2)
```

################################ 

Instead of using an Rproject this week we are getting to manually set our working directory (Translation: File where everything will be saved.)Always set your wd before working in R. This ensures that all your work(graphics, subsets, etc) will be saved in the same place and that R will know where to look for your data.

After the setwd() command insert the file path to your own folder that you will use for this class(If it doesn't exist create it using finder or in windows).Make sure both files I emailed you, this .rmd file and the .rdata file are in that folder.

There are two ways to have R tell you the file path. First, you can use the file.choose() to find the file path.Just copy and paste from the console into this code. The other is to load or set the wd directly in Rstudio using the file pane to the right, but if you do this you need paste into your code so it will knit.

You must have the setwd() command in your code to knit. 

This is perhaps the most counter-intuitive thing you'll do in R, but it is important. Remember, do not use point and click, you need the code to knit.

```{r}
getwd() #Returns which directory you are currently working in
setwd("~/Dropbox/POLS 641/CMU Su26/cmucourse26/ClassCode/Day 3 - ") #Sets working directory - You will need to copy and paste your file path after using file.choose() in the console below. 
getwd() #make sure using right folder
```

We're going to use data from the article "Facing Change: Gender and Climate Change Attitudes Worldwide" by Bush and Clayton Data can be found here: <https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/0Y8IKW>

This is just one of the many data sources they use.

```{r}
pewdata <- read.csv("pewdata.csv") #If working directory is set correctly and this file is in it this command should work. 
```

Lets explore the data: Variable names, \# of variables, \# of respondents

```{r}
dim(pewdata)
names(pewdata)

```

Thats a lot of respondents.

```{r}
table(pewdata$Country)
length(unique(pewdata$Country))

#How many respondents from Thailand?

length(unique(pewdata$WorryType))

prop.table(table(pewdata$Country))

str(pewdata) #This tells us what type of data R thinks we have
```

Look at climate concern

```{r}
mean(pewdata$ClimConcen)#Doesn't know what to do with NAs
mean(pewdata$ClimConcen, na.rm = T)
median(pewdata$ClimConce, na.rm = T)
summary(pewdata$ClimConcen)

hist(pewdata$ClimConcen)

#A good way to count NAs in a column
sum(is.na(pewdata$ClimConcen))
sum(is.na(pewdata$Ideo))

table(pewdata$Country, pewdata$Ideo)
```

Sex differences

```{r}
males <- pewdata[which(pewdata$Female==0),]
females <- pewdata[which(pewdata$Female==1),]

summary(males$ClimConcen)
summary(females$ClimConcen)
```

Lets say we we were interested in Income and climate concern . . . .

Use base R to create scatterplot

```{r}
plot(pewdata$StandIncome, pewdata$ClimConcen) #Not very helpful
```

Reduce N to make more useful

```{r}
pewsamp1 <- pewdata[sample(nrow(pewdata), 500), ]# a random sample of 500 ie a new subset

mean(pewsamp1$ClimConcen, na.rm=T)
mean(pewdata$ClimConcen, na.rm=T)

plot(pewsamp1$StandIncome, pewsamp1$ClimConcen, main="Income and Climate Concern", xlab="Income", ylab="Climate Concern") #our standard scatter plot
abline(lm(pewsamp1$ClimConcen~pewsamp1$StandIncome))
```

Using tidyverse

```{r}
pewsamp2 <- sample_n(pewdata, 500)#Notice how tidyverse uses a slighty different vocabulary and grammar

mean(pewsamp2$ClimConcen, na.rm=T)#Ack! This is different from above

#We can create the same figure with the ggplot package. ggplot uses a different grammar.
ggplot(data=pewsamp2, aes(x=StandIncome, y=ClimConcen))+
  geom_point(shape=1)+
  xlab("Income")+ ylab("Climate Concern")+geom_smooth(method='lm', formula= y~x)
```

There are many ways to code in R. Base R can do everything you need, but I find "the tidyverse" more intuitive so I am introducing it early with the assumption you'll find it easier as well.

This is interesting, but might want to look at one country at a time.

Subset the data

```{r}
usa <- pewdata[which(pewdata$Country=="United States"),]
indo <- pewdata[which(pewdata$Country=="Indonesia"),]

plot(usa$StandIncome, usa$ClimConcen, main="Income and Climate Concern", xlab="Income", ylab="Climate Concern") #our standard scatter plot
abline(lm(usa$ClimConcen~usa$StandIncome), col="red")
abline(lm(pewdata$ClimConcen~pewdata$StandIncome), col="blue")#add line for whole sample to make comparison
abline(lm(indo$ClimConcen~indo$StandIncome), col="pink")
```

Create same graph using dplyr and ggplot

```{r}
USA <- filter(pewdata, Country=="United States") #This is dplyr version of what we did last time. Check out dplyr cheat sheet posted to dropbox

ggplot(USA, aes(x=StandIncome, y=ClimConcen))+
  geom_point(shape=1)+
  xlab("Income")+ ylab("Climate Concern")+geom_smooth(method='lm', formula= y~x, col="red")

ggplot(USA, aes(x=StandIncome, y=ClimConcen))+
  geom_point()+ ggtitle("United States")+
  xlab("Income")+ ylab("Climate Concern")+geom_smooth(method='lm', formula= y~x, col="red", se=F)+geom_jitter() + theme_test()
```

An especially nice feature of tidy verse is that you can use piping(%\>%) to do multiple things in a single line of code

```{r}
pewdata %>% filter(Country=="Russia") %>%  ggplot(aes(x=StandIncome, y=ClimConcen))+
  geom_point()+ ggtitle("Russia")+
  xlab("Income")+ ylab("Climate Concern")+geom_smooth(method='lm', formula= y~x, col="red")+geom_jitter() +theme_bw()

#This can make it much easier to run similar versions of the same graph
pewdata %>% filter(Country=="Australia") %>%  ggplot(aes(x=StandIncome, y=ClimConcen))+
  geom_point()+ ggtitle("Australia")+
  xlab("Income")+ ylab("Climate Concern")+geom_smooth(method='lm', formula= y~x, col="red")+geom_jitter() +theme_bw()

pewdata %>% filter(Country=="France") %>%  ggplot(aes(x=StandIncome, y=ClimConcen))+
  geom_point()+ ggtitle("France")+
  xlab("Income")+ ylab("Climate Concern")+geom_smooth(method='lm', formula= y~x, col="red")+geom_jitter() +theme_bw()

pewdata %>% filter(Female==1, Country=="France") %>% ggplot(aes(x=StandIncome, y=ClimConcen))+
  geom_point()+ ggtitle("Females")+
  xlab("Income")+ ylab("Climate Concern")+geom_smooth(method='lm', formula= y~x, col="red")+geom_jitter() +theme_bw()
```

Your turn to play - create for another county, or look at another DV

```{r}


```

Recreating Figure 1

```{r}
ggplot(pewdata, aes(x=GDPPC)) + 
  geom_density() #ggplot code for a histogram

pewdata$female <- ifelse(pewdata$Female==1, "female", "male")

bycountryg <- pewdata %>% group_by(Country, female) %>% summarise(Avgserios=mean(ClimSerious, na.rm=T), GDP=mean(GDPPC))%>% pivot_wider(names_from = female, values_from = Avgserios)%>% mutate(diff=female-male)
bycountryg


ggplot(bycountryg, aes(x=log(GDP), y=diff, label=Country))+geom_point()+geom_text()+ ggtitle("Gender Gap by GDP")+
  xlab("GDP per Capita(logged)")+ ylab("Gender Gap")+geom_hline(yintercept=0, col="lightblue")+ geom_smooth(method='lm', formula= y~x, col="coral", se=F)
```

Make sure to knit and, if you'd like, save the data.
